---
title: "Alternative Teaching Method Sample"
author: "Gabriela Grasmane"
date: "`r Sys.Date()`"
output: rmarkdown::github_document
---

# Introduction

This R Notebook provides an example of the alternative teaching method that I envision to apply in the design of my workshop for SCICOGN301, *Advanced Cognitive Science*.

------------------------------------------------------------------------

**Data:** `nycflights13::flights`

**Research Question:** *Which NYC airport has the biggest average departure delay?*

------------------------------------------------------------------------

This sample has the following simple structure:

1.  Small example of a function
2.  Application of the example
3.  Repeat 1 and 2 for all required functions
4.  Result is achieved step by step

## Setup

To run the following code block, press the green button `Run Current Chunk` on the top right corner of the block.

```{r}
install.packages("tidyverse")
library(tidyverse)
library(nycflights13)
```

-   `tidyverse` is a collection of R packages designed following tidy principles for readable code. Most functions that we use for basic data analysis come from this package.
-   `nycflights13` is a package that contains data on all flights departing from New York City (JFK, LGA or EWR airports) in 2013. We will use the `flights` data from this package to answer our research question.

------------------------------------------------------------------------

# Dataset

The following code block assigns the flights data from the `nycflights13` package to your `data` variable. You can try to rename the variable to your liking.

```{r}
# Load the dataset
data <- flights
```

## Viewing the dataset in RStudio

Before proceeding, I want to tell you about the **pipe operator**. It looks like this: `|>`. Its keyboard shortcut is `Ctrl (or Cmd) + Shift + M`.

The pipe operator is used to pass the result of one step directly into the next step. It is convenient because it allows to create a tidy sequence of functions.

To try it, use the function `view()` to look at the dataset. Keep in mind that `data` precedes the pipe.

```{r}
# Substitute this comment with code (you may give it a few seconds to run)
```

Look at the variables in the dataset: which two variables do we need to answer our research question?

------------------------------------------------------------------------

# Tools

## 1. `filter()`

Run the following code. What does the `filter()` command do?

```{r}
data |> 
  filter(origin == "JFK") |> 
  view()
```

If we investigate the distinct observations in the `origin` variable, we will see that flights depart from three airports:

```{r}
data |> 
  distinct(origin)
```

However, the `dep_delay()` variable contains missing values.

```{r}
data |> 
  count(is.na(dep_delay))
```

-   `count()` counts the number of observations in a variable.
-   `is.na()` outputs a `TRUE` or `FALSE` value, which is determined by whether an observation is missing (**n**ot **a**vailable).

You need to remove these missing values to calculate the average departure delay. This can be done by using the `filter()` function and the negation operator `!` at the beginning inside the parentheses (as we say, at the beginning of the *argument*). Let's start looking for the most delayed airport.

```{r}
most_delayed_airp <- 
  data |> 
  # Filter the dep_delay values that are not NA
```

Are there any NA values left?

```{r}
most_delayed_airp |> 
  count(is.na(dep_delay))
```

## 2. `group_by()` and `summarize()`

These two functions are generally used together, as they analyze data in **groups**. Do you understand why it is useful to group a variable in our case?

~~Hint: remember the distinct values of the `origin` variable.~~

-   `group_by()` divides the data into smaller groups;
-   `summarize()` creates a new variable that calculates something for each group.

Inside the `summarize()` function, we want to create an average departure delay variable called `avg_dep_delay`. It is computed for each airport by using the `mean()` function. Which variable are we using to calculate the mean of the departure delay?

```{r}
most_delayed_airpo <- 
  most_delayed_airp |> 
  # Substitute this comment with the function that will group the data
  # Uncomment the following two lines of code and complete it:
  # summarize(avg_dep_delay = mean(dep_delay)) |> 
  # print()
```

## 3. `arrange()`

This function sorts the data values, either in ascending or descending order. You should specify one or the other with the `asc()` or `desc()` functions respectively to arrange the values of `avg_dep_delay`.

```{r}
most_delayed_airpor <- 
  most_delayed_airpo |> 
  # Arrange the result
  # print()
```

------------------------------------------------------------------------

Until now, I tried to explain some functions and common syntax to warm you up to what we are doing here. However, it is good practice to write code in one pipeline, rather than in separate blocks. Below you can see a tidy version for the procedure we just went through to answer our research question:

```{r}
most_delayed_airport <- 
  data |> 
    filter(!is.na(dep_delay)) |> 
    group_by(origin) |> 
    summarize(avg_dep_delay = mean(dep_delay)) |> 
    arrange(desc(avg_dep_delay))
```

You can see the result by running the following block.

```{r}
most_delayed_airport
```
