---
title: "Alternative Teaching Method Sample"
author: "Gabriela Grasmane"
date: "`r Sys.Date()`"
output: 
  html_notebook: # html_document if swithing to Rmd
    theme: default
---

# Introduction

Welcome to this beginner-friendly R tutorial! In this guide, we will use R to answer a simple research question. No prior programming experience is required. By the end of this notebook, you will learn how to:

- Load and explore a dataset.
- Perform basic data analysis.
- Visualize your results.

## Research Question

**Example:** What is the average age of participants in a sample dataset?

---

## Step 1: Setting Up Your Environment

Before we begin, ensure you have R and RStudio installed on your computer. If not, download them from the following links:

- [R](https://www.r-project.org/)
- [RStudio](https://rstudio.com/)

### Loading Required Libraries

Libraries in R are like toolkits that extend its functionality. Let’s load the necessary libraries.

```{r}
# Install libraries if you haven't already
if (!require("tidyverse")) install.packages("tidyverse")

# Load the tidyverse package
library(tidyverse)
```

---

## Step 2: Loading the Dataset

For this tutorial, we will use a sample dataset. If you have your own dataset, you can load it using the same steps.

### Example Dataset

We will use the built-in `mtcars` dataset, which contains information about cars, such as their weight, miles per gallon (mpg), and number of cylinders.

```{r}
# Load the dataset
my_data <- mtcars

# View the first few rows of the dataset
head(my_data)
```

### Explanation
- **`my_data <- mtcars`**: Assigns the `mtcars` dataset to a new variable `my_data`.
- **`head(my_data)`**: Displays the first 6 rows of the dataset.

---

## Step 3: Exploring the Data

It’s important to understand the structure and contents of the dataset before analyzing it.

```{r}
# Check the structure of the dataset
str(my_data)

# Get a summary of the dataset
summary(my_data)
```

### Explanation
- **`str(my_data)`**: Shows the structure of the dataset, including variable types.
- **`summary(my_data)`**: Provides basic statistical summaries for each variable.

---

## Step 4: Analyzing the Data

Now, let’s answer our research question: *What is the average mpg of the cars in the dataset?*

```{r}
# Calculate the average mpg
average_mpg <- mean(my_data$mpg)

# Print the result
average_mpg
```

### Explanation
- **`my_data$mpg`**: Accesses the `mpg` column in the dataset.
- **`mean()`**: Calculates the average of a numeric vector.

---

## Step 5: Visualizing the Data

Visualizations help us better understand the data. Let’s create a histogram of the `mpg` variable.

```{r}
# Create a histogram of mpg
my_data %>%
  ggplot(aes(x = mpg)) +
  geom_histogram(binwidth = 2, fill = "blue", color = "black") +
  labs(title = "Distribution of MPG", x = "Miles Per Gallon (MPG)", y = "Count")
```

### Explanation
- **`ggplot()`**: A function from the `ggplot2` package for creating visualizations.
- **`aes(x = mpg)`**: Specifies that the x-axis will represent the `mpg` variable.
- **`geom_histogram()`**: Creates a histogram with specified bin width and colors.

---

## Step 6: Summarizing Your Findings

Based on our analysis, the average `mpg` of the cars in this dataset is approximately `r round(average_mpg, 2)`. The histogram shows how the `mpg` values are distributed.